---
title: "R-ticulate Exercises"
author: "Martin Bader and Sebastian Leuzinger"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  Some of What Lies Ahead
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
## List of packages required for this tutorial
# pkgs <- c("learnr", "ggplot2", "nlsMicrobio", "mgcv")

# ## Check whether the specified package is available
#   inst.pack <- rownames(installed.packages()) # list of installed packages
#   if (length(setdiff(x = pkgs, y = inst.pack)) > 0) { # setdiff() returns the elements of x that don't occur in y
#     install.packages(setdiff(pkgs, inst.pack), repos = "https://cloud.r-project.org", dependencies = T)  
#   }

## Load required packages
library(learnr)
library(nlsMicrobio)
library(mgcv)
library(ggplot2)

## Hailstorm probability as a function of supercooled water droplet concentration in the atmosphere ----
# Set seed for reproducibility
set.seed(789)

# Number of data points
n <- 1000

# Generate random data for supercooled water droplets (values between 0 and 2 g/m³)
supercooled_water_droplets <- runif(n, min = 0, max = 2)

# Define logistic regression model coefficients
# Logistic regression model: log(p / (1 - p)) = beta0 + beta1 * supercooled_water_droplets
# Plausible coefficients for hailstorm prediction
beta0 <- -4     # Intercept (base level of hailstorm occurrence)
beta1 <- 3      # Coefficient for supercooled water droplets (impact on hailstorm occurrence)

# Calculate the probability of a hailstorm based on supercooled water droplet concentration
# Logistic function:
prob_hailstorm <- 1 / (1 + exp(-(beta0 + beta1 * supercooled_water_droplets)))

# Generate the binary outcome (1 = hailstorm, 0 = no hailstorm)
hailstorm <- rbinom(n, size = 1, prob = prob_hailstorm)

# Create a dataframe with the simulated data
hail <- data.frame(
  scwd = round(supercooled_water_droplets, digits = 2),
  hailstorm = hailstorm
)


## Crossbills ----
# crossbills <- data.frame(direction = factor(c(rep(c("left", "right"), times = c(100, 311)), rep(c("left", "right"), times = c(90, 198)))), gender = factor(rep(c("male", "female"), times = c(411, 288))))

# m1 <- glm(direction ~ gender, data = crossbills, family = binomial)
# m1 <- glm(direction ~ 1, data = crossbills, family = binomial)
# summary(m1)

# group_by(crossbills, direction) %>% summarise(n = n()) %>% mutate(freq = n/sum(n))
# crossbills2 <- group_by(crossbills, gender, direction) %>% summarise(n = n()) %>% mutate(freq = n/sum(n))
# newdat <- data.frame(gender = levels(crossbills$gender))
# preds <- predict(m1, newdata = newdat, se = T)

# ggplot(data = crossbills2, aes(x = direction, y = n)) +
  # geom_bar(stat = "identity") +
  # facet_wrap(facets = ~ gender)
```

## Chapter 12 - Some of What Lies Ahead

### Exercise 1 - Multiple-choice quiz

Please note that there is only one correct (or most correct) answer to each question.

<p style="margin-top:-30px;">
</p>
```{r anova_quiz, echo=FALSE}
quiz(caption = "",
      question("Generalised linear models (GLMs) relax the assumption of",
         answer("independence among observations."),
         answer("variance homogeneity (homoscedasticity)."),
         answer("normally distributed model residuals (errors).", correct = T),
         answer("a constant slope."),
         answer("None of the above."),
         allow_retry = T
         ),
     
     question("The most appropriate error distribution for a binary response variable such as dead/alive or presence/absence, is the",
         answer("gaussian (normal) distribution."),
         answer("binomial distribution.", correct = T),
         answer("poisson distribution."),
         answer("gamma distribution."),
         answer("None of the above."),
         allow_retry = T
         ),
     
      question("What is the purpose of the link function in generalised linear models?",
         answer("To transform the response variable so that it follows a normal distribution."),
         answer("To ensure the model residuals (errors) are homoscedastic."),
         answer("To provide a non-linear relationship between the predictors and the response variable."),
         answer("To relate the linear predictor to the expected values of the response variable.", correct = T),
         answer("None of the above."),
         allow_retry = T
         ),
     
         question("In a figure related to a logistic regression model (binomial GLM) you notice symmetric confidence intervals around the fitted curve. What do you conclude?", 
         answer("This is the way it should be suggesting an accurate visualisation of the model results."),
         answer("Something is wrong. It is not possible to calculate confidence intervals for GLMs."),
          answer("This indicates that an ill-suited distribution was chosen."),
          answer("The confidence interval was incorrectly calculated after back-transforming the predictions from the link scale.", correct = T),
         answer("None of the above."),
         allow_retry = T
         ),
     
     question("One of the basic requirements of a nonlinear regression model is",
         answer("a nonlinear equation describing the relationship between response and predictor.", correct = T),
         answer("an intercept and slope."),
         answer("a distribution for the response variable and model errors."),
         answer("a correct link function."),
         answer("None of the above."),
         allow_retry = T
         ),
     
         question("What does the term 'starting values' mean in the context of nonlinear regression?",
         answer("This refers to the first few rows in the dataset containing the nonlinear relationship."),
         answer("This refers to the starting points of the desired range of the continuous predictor."),
         answer("These are initial parameter estimates that need to be supplied to provide starting points for the optimisation algorithm of the model.", correct = T),
         answer("This refers to the starting points of the range of the model predictions, which allows the user to specify predictions for interpolation or extrapolation."),
         answer("This refers to start values that need to be provided to estimate the confidence interval around the nonlinear curve."),
         allow_retry = T
         ),
     
      question("Which of the following statements is true for generalised additive models (GAMs)?",
         answer("GAMs assume normally distributed errors (residuals)."),
         answer("In GAMs, the relationship between the response variable and each predictor is assumed to be linear."),
         answer("GAMs allow for the inclusion of non-linear relationships between the predictors and the response variable by using smooth functions.", correct = T),
         answer("The interpretation of coefficients in a GAM is the same as in a linear regression model."),
         answer("GAMs require starting values as starting points for the fitting algorithm similar to nonlinear regression models."),
         allow_retry = T
         ),
     
      question("The summary output of generalised additive models",
         answer("is the same as in nonlinear models."),
         answer("consists of two parts, a parametric component and a smoother-related component.", correct = T),
         answer("only consists of the estimated degrees of freedom of the smoother, indicating the degree of nonlinearity, and a goodness-of-fit statistic."),
         answer("consists of an intercept and the associated standard error, _t_- and _P_-value."),
         answer("None of the above."),
         allow_retry = T
         ),
     
        question("Which of the following statements best describes heteroscedasticty (variance heterogeneity)?",
        answer("The residuals are normally distributed."),
        answer("The variance of the residuals is constant across all levels of the independent variables."),
        answer("The variance of the residuals increases or decreases with the level of the fitted values or the independent variable(s), respectively.", correct = T),
        answer("The residuals are independent and identically distributed."),
        answer("None of the above."),
  allow_retry = T
),

question("Which of the following methods can be used to detect heteroscedasticity in a regression model?",
        answer("Shapiro-Wilk test."),
        answer("Qantile-quantile plot."),
        answer("Variance inflation factor (VIF)."),
        answer("Histogram"),
        answer("None of the above.", correct = T),
  allow_retry = T
),

question("Why is it important to check for and deal with heteroscedasticity if it occurs?",
        answer("Heteroscedasticity commonly results in biased estimates of the standard errors of the regression coefficients and thus compromised statistical inference.", correct = T),
        answer("Since heteroscedasticity can lead to biased estimates of the regression coefficients."),
        answer("Since heteroscedasticity indicates that the linear relationship between the predictors and the response variable is incorrect, requiring a respecification of the model."),
        answer("Heteroscedasticity results in the loss of observations and thus reduces the sample size."),
        answer("None of the above."),
  allow_retry = T
),

question("Which of the following techniques is not a suitable remedy for heterscedasticity?",
        answer("A variance-stabilising data transformation."),
        answer("Using robust, heteroscedasticity-consistent estimation of standard errors."),
        answer("Using a generalised linear model framework, allowing the use of various error distributions.", correct = T),
        answer("Using a generalised least squares approach, which explicitly models the variance structure of the errors and adjusts the estimation process accordingly.
"),
        answer("None of the above."),
  allow_retry = T
)
)
```

The model code related to the exercises below relies on the following R packages (which are autoloaded behind the scenes in this tutorial):

* _ggplot2_
* _nlsMicrobio_
* _mgcv_

### Exercise 2 - Logistic Regression

Supercooling refers to the process of lowering the temperature of a liquid below its normal freezing point without transition into the solid phase. Supercooled water droplets are common in the atmosphere, where their availability is a driver for hail formation. Here, we will analyse the occurrence of hailstorms as a function of the concentration of supercooled water in the atmosphere using the pre-loaded dataset `hail`, containing the supercooled water droplet concentration in the `scwd` column and the occurrence of hailstorms given as a binary variable in the `hailstorm` column.

a. Visualise the data

b. Run a generalised linear model with a binomial error distribution and the default logit link.

c. Calculate the model predictions including a 95% confidence interval, back-transform them to the response scale and add them to the previously created plot.

d. At roughly what supercooled water droplet concentration is the hailstorm probability 50%?

```{r hail-exercise, exercise=TRUE, fig.align='center', fig.width=4.5, fig.height=4.5, message=FALSE}
## View the first few observations
head(hail)
```
```{r hail-exercise-solution, exercise.reveal_solution = TRUE}
## a. Visualise the data
ggplot(data = hail, aes(x = scwd, y = hailstorm)) +
  geom_point(alpha = 0.3, position = position_jitter(width = 0, height = 0.05)) +
  labs(x = expression(paste("Supercooled Water Droplets (g m"^-3, ")")),
       y = "Hailstorm probability") +
  theme_test()

## b. Run a binomial GLM
m1 <- glm(hailstorm ~ scwd, data = hail, family = binomial()) # binomial() defaults to binomial(link = "logit")
summary(m1)

## c. Model predictions incl. 95% confidence interval
## Set up a new data frame with a fine grid of scwd values to predict from (ensures a smooth regression curve)
newdat <- data.frame(scwd = seq(min(hail$scwd), max(hail$scwd), length.out = 200))
preds <- predict(m1, newdata = newdat, se = T)

bt <- binomial()$linkinv # back-transformation function, see help file for 'family'
newdat$fit <- bt(preds$fit)
newdat$upper <- bt(preds$fit + 1.96 * preds$se.fit)
newdat$lower <- bt(preds$fit - 1.96 * preds$se.fit)
newdat$hailstorm <- 0 # add 'empty' hailstorm variable, required for plotting (ggplot assumes this variable as it occurrs in the hail data frame)

# Plot to visualise the relationship
ggplot(data = hail, aes(x = scwd, y = hailstorm)) +
  geom_ribbon(data = newdat, aes(x = scwd, ymin = lower, ymax = upper), fill = "grey50", alpha = 0.5) +
  geom_line(data = newdat, aes(x = scwd, y = fit), linewidth = 1, colour = "red3") +
  geom_point(alpha = 0.3, position = position_jitter(width = 0, height = 0.05)) +
  labs(x = expression(paste("Supercooled Water Droplets (g m"^-3, ")")),
       y = "Hailstorm probability") +
  theme_test()
```
</details>

<p style="margin-top:40px">
<details><summary>Click <a class="hover" text-decoration="underline" style="color:blue;cursor:pointer">here</a> to view the solution to question 2 d.</summary>
<p style="margin-top:10px">
The hailstorm probability reaches about 50% at a supercooled water droplet concentration of _ca._ 1.4 g m^-3^.
</p></details></p>

### Exercise 3 - Building a Nonlinear Regression Model

The data set `L.minor` (package _nlsMicrobio_) contains nitrogen uptake rates (variable `rate`) as a function of aquatic nitrogen concentration (variable `conc`). The data stem from Cedergreen & Madsen (2002) Nitrogen uptake by the floating macrophyte _Lemna minor_. _New Phytologist_, 155, 285–292.

Uptake kinetics are often described by the Michaelis-Menten model:

$$
f(x) = \frac{V_{max}\: x}{K_{m} + x}
$$
Here, $f(x)$ is the nitrogen uptake rate, $x$ is the substrate concentration, _V_~max~ is the maximum reaction rate under substrate-saturated conditions and _K_~m~ indicates the Michaelis constant, which is the substrate concentration at which the reaction rate is half of _V_~max~. It reflects the affinity of the enzyme for the substrate; a lower _K_~m~ indicates higher affinity.

a. Plot the data. 

b. Guess intital parameter estimates (eyeball the asymptote and the value on the _x_-axis, where the rate is about half of the max) and fit the Michaelis-Menten model model to the data. 

c. Check the underlying model assumptions of normality and variance homogeneity of the residuals. 

d. Calculate the model predictions and add the regression curve to the plot. 


```{r nls-exercise, exercise=TRUE, fig.align='center', fig.width=4.5, fig.height=4.5, message=FALSE}
## Load the data
data(L.minor)

## View the first few observations
head(L.minor)
```

```{r nls-exercise-solution, exercise.reveal_solution = TRUE}
## a. Plot the data
plot(rate ~ conc, data = L.minor, ylim = c(0, 120))

## b. Run the nonlinear model
m1 <- nls(rate ~ Vm * conc/(K + conc), data = L.minor, start = c(Vm = 110, K = 20))
summary(m1)

## c. Check model assumptions
qqnorm(resid(m1))
qqline(resid(m1))
plot(resid(m1) ~ fitted(m1))
abline(h = 0, lty = 2)

## d. Add regression curve
plot(rate ~ conc, data = L.minor, ylim = c(0, 120))
lines(L.minor$conc, fitted(m1)) # kinked looking curve because there is no interpolation between measured concentrations

## Set up a new data frame containing a fine grid of predictor values( x-values) to predict from in order to obtain a smooth regression curve. NOTE that the variable name in the new data frame needs to match the name in the nonlinear model exactly (here 'conc').
newdat <- data.frame(conc = seq(min(L.minor$conc), max(L.minor$conc), length.out = 200))
newdat$fit <- predict(m1, newdata = newdat)
plot(rate ~ conc, data = L.minor, ylim = c(0, 120))
lines(fit ~ conc, data = newdat)
```

### Exercise 4 - Generalised additive modeling

We will revert to the built-in `airquality` dataset and model ozone concentration (variable `Ozone`) as a function of solar radiation (variable `Solar.R`) using a generalised additive model (GAM). As we have determined in chapter 3, the `Ozone` variable follows a gamma distribution, so you need to specify this distribution via the `family` argument (note that the function name for the gamma distribution is spelled with a capital 'G', see help file for `family`). Use the default "inverse" link of the gamma distribution.

a. Create an exploratory plot of the data.

b. Fit a GAM with gamma error distribution.

c. Check the model assumptions using the `gam.check` function.

d. Calculate the model predictions and add the GAM fit including its 95% confidence interval to the plot.

**Hint:** Use `par(mfrow = c(2, 2)` to split the plotting area in four panels before using the `gam.check` function to check the model assumptions.

```{r gam-exercise2, exercise=TRUE, exercise.eval=TRUE, fig.align='center', fig.width=4.5, fig.height=4.5, message=FALSE}
## Load the data
data(airquality)

## View the first few observations
head(airquality)
```
```{r gam-exercise2-solution, exercise.reveal_solution = TRUE, message=FALSE}
## a. Exploratory plot 
plot(Ozone ~ Solar.R, data = airquality)

## b. GAM with gamma error distribution
m1 <- gam(Ozone ~ s(Solar.R), data = airquality, family = Gamma()) # Gamma() boils down to Gamma(link = "inverse")
summary(m1)

## c. Check model assumptions
op <- par(mfrow = c(2, 2))
gam.check(m1)
par(op) # reset the graphics defaults

## d. Model predictions
newdat <- with(airquality, data.frame(Solar.R = seq(min(Solar.R, na.rm = T), max(Solar.R, na.rm = T), length.out = 200)))
preds <- predict(m1, newdata = newdat, se.fit = T)
newdat$fit <- 1/preds$fit # 1/preds$fit is the back-transformation of the inverse link. We could have created a function like this: bt <- Gamma()$linkinv, but this may seems overkill for such a simple back-transformation
newdat$upper <- 1/(preds$fit + 1.96 * preds$se.fit) # recall that the calculation of the confidence interval must be done on the link scale, prior to the back-transformation to the response scale
newdat$lower <- 1/(preds$fit - 1.96 * preds$se.fit)

plot(Ozone ~ Solar.R, data = airquality, las = 1, ylab = "Ozone (ppb)", xlab = " Solar radiation (Ly)") # Ly = langley, a unit of heat transmission
with(newdat, polygon(x = c(Solar.R, rev(Solar.R)), y = c(upper, rev(lower)), col = "grey", border = "grey"))
points(Ozone ~ Solar.R, data = airquality)
lines(fit ~ Solar.R, data = newdat)
```
