---
title: "R-ticulate Exercises"
author: "Martin Bader and Sebastian Leuzinger"
output: learnr::tutorial
runtime: shiny_prerendered
description: >
  Communicating Quantitative Information Using Visuals
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
## List of packages required for this tutorial
# pkgs <- c("learnr", "dplyr", "ggplot2", "ggpubr", "cowplot")
# 
# ## Check whether the specified package is available
#   inst.pack <- rownames(installed.packages()) # list of installed packages
#   if (length(setdiff(x = pkgs, y = inst.pack)) > 0) { # setdiff() returns the elements of x that don't occur in y
#     install.packages(setdiff(pkgs, inst.pack), repos = "https://cloud.r-project.org", dependencies = T)  
#   }
  
## Load required packages
library(learnr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(cowplot)

# Grouped boxplot example
# ss <- sample(1:10000, 1) # 2007, 5776, 5120
set.seed(5776)
site <- rep(LETTERS[1:7], each = 20)
treat <- rep(c("exclusion","control"), each = 10)
length <- sample(seq(21, 39, by = 0.5), size = 140, replace = T) + sample(x = 1:3, size = 140, replace = T)
insects <- data.frame(site = factor(site), treat = factor(treat), length)
insects <- mutate(.data = insects, length = case_when(treat == "exclusion" ~ length + 2,
                                treat == "control" ~ length - 2))
```

## Chapter 06 - Communicating Quantitative Information Using Visuals

### Exercise 1 - Multiple-choice quiz

Please note that there is only one correct (or most correct) answer to each question.

<p style="margin-top:-30px;">
</p>

```{r anova_quiz, echo=FALSE}
quiz(caption = "",
       question("Which of the below does not represent a purpose of statistical graphs?",
        answer("To give a description of the observations.", correct = T),
        answer("To identify patterns in data."),
        answer("To summarise data."),
        answer("To compare data."),
        answer("None of the above."),
  allow_retry = T
),
question("If you have a continuous response variable and a categorical explanatory variable, then the most appropriate visualisation can be achieved with",
        answer("Scatter plots and dotcharts."),
        answer("Histograms and quantile-quantile plots."),
        answer("Mosaic plots and piecharts."),
        answer("Boxplots and bar charts.", correct = T),
        answer("None of the above."),
  allow_retry = T
),
      question("Which function is used to create a basic scatter plot in R?",
         answer("`plot`", correct = T),
         answer("`scatter`"),
         answer("`barplot`"),
         answer("`hist`"),
         answer("`boxplot`"),
         allow_retry = T
         ),
     
     question("A boxplot is often preferable to a barplot because",
         answer("it is easier to render for the graphics software and may use less ink in print."),
         answer("it shows the distribution of the data with a five number summary (minimum, lower quartile, median, upper quartile, maximum), while a barplot only shows the mean and some measure of dispersion with the error bars.", correct = T),
         answer("it can more effectively display the mean of a particular dataset."),
         answer("it highlights categorical data more effectively."),
         answer("it requires less data to produce an informative visualisation including the extremes, median, quartiles, and potential outliers."),
         allow_retry = T
         ),
     
      question("Which function is used to save a plot as a PNG file in R?",
         answer("`dev.print`"),
         answer("`writePlot`"),
         answer("`writePNG`"),
         answer("`tiff`"),
         answer("`png`", correct = T),
         allow_retry = T
         ),
     
         question("Which function is used to change the overall appearance of a plot created with the _ggplot2_ package?", 
         answer("`ggplot`"),
         answer("`qplot`"),
         answer("`theme`", correct = T),
         answer("`element_blank`"),
         answer("The default settings cannot be changed."),
         allow_retry = T
         ),
     
     question("Which of the following sets contains only high-level plotting functions in base R?",
         answer("`plot`, `hist`, `barplot`, `points`, `pairs`"),
         answer("`image`, `boxplot`, `hist`, `qqnorm`, `plot`", correct = T),
         answer("`coplot`, `barplot`, `hist`, `pairs`, `polygon`"),
         answer("`hist`, `box`, `barplot`, `qqnorm`"),
         answer("None of the above."),
         allow_retry = T
         ),
     
         question("What is a box-and-violin plot?",
         answer("A combination of a histogram and a scatter plot to illustrate data distribution and individual data points."),
         answer("A combination of a bar chart and a scatter plot to show both categorical and continuous data."),
         answer("A combination of a line chart and a pie chart to show changes in data over time and proportions."),
         answer("A combination of a boxplot and an upended, mirrored density histogram.", correct = T),
         answer("A heatmap-style visualisation with inset boxes to indicate areas of interest."),
         allow_retry = T
         ),
     
      question("What does the function `facet_wrap` do in R package _ggplot2_?",
         answer("Creates a multi-panel figure by arranging individual plots in a grid layout based on a categorical variable.", correct = T),
         answer("Adds a background grid to the plot."),
         answer("Adds a smoothing line to the data points in the plot."),
         answer("Scales the x-axis based on the data range without lower or upper offsets."),
         answer("Scales x- and y-axes based on the data range without lower or upper offsets."),
         allow_retry = T
         ),
     
      question("Colour contour plots are an excellent tool for",
         answer("colour-coding values of a discrete variable arranged in a grid."),
         answer("visualising continuous 3D data in a 2D graph.", correct = T),
         answer("creating 3D visualisations and animations."),
         answer("creating 3D scatterplots."),
         answer("None of the above."),
         allow_retry = T
         ),

question("To quickly visualise all possible bivariate relationships between multiple continuous variables in a dataset, we can use the functions",
        answer("`pairs` (base R) or `ggpairs` (_ggplot2_ framework).", correct = T),
        answer("`layout` (base R) or `facet_wrap` (_ggplot2_ framework)."),
        answer("`coplot` (base R) or `facet_wrap` (_ggplot2_ framework)."),
        answer("`boxplot` (base R) or `geom_boxplot` (_ggplot2_ framework)."),
        answer("None of the above."),
  allow_retry = T
),

question("Special formatting is often needed for printing units with subscript, superscript, Greek letters etc. Which of the following combinations of functions is commonly used to create scientific units?",
        answer('`deparse(substitute(...))`'),
        answer('`eval(parse(...))`'),
        answer('`substitute(expression(...))`'),
        answer('`expression(paste(...))`', correct = T),
        answer("None of the above."),
  allow_retry = T
)
)
```

The model code related to the exercises below relies on the following R packages (which are autoloaded behind the scenes in this tutorial):

* _dplyr_
* _ggplot2_
* _ggpubr_
* _cowplot_

### Exercise 2 - Exploratory Plots

2.1 Create a vector of 100 random variables drawn from a standard normal distribution (_i.e._ mean = 0, standard deviation = 1) using the `rnorm` function. Draw a histogram overlaid with a density curve using both a base R and a _ggplot2_ approach (see help for `geom_histogram` and `geom_density`).

```{r hist-exercise, exercise=TRUE, exercise.eval=TRUE, fig.align='center', fig.width=4.5, fig.height=4.5, message=FALSE, warning=FALSE}

```

```{r hist-exercise-solution, exercise.reveal_solution = TRUE}
## Histogram with density curve
## Create a vector of 100 random numbers from a standard normal distribution
random <- data.frame(rn = rnorm(100))

## Base R
hist(random$rn, main = NA, freq = F) # freq = F makes sure that the probability density instead of the raw counts are used
lines(density(random$rn))

## gglot2
ggplot(data = random, aes(x = rn)) +
  geom_histogram(aes(y = after_stat(density))) +
  geom_density()

## Alternative using the gghistogram function in package ggpubr
gghistogram(random, x = "rn", y = "density", add_density = T, fill = "grey50", color = "grey50")
```
<br>

2.2 Create a boxplot of the built-in dataset `Loblolly` (growth of Loblolly pine trees), using the response variable `height` as a function of `Seed` (seed source of the trees). Use both base R and _ggplot2_ for this exercise.

```{r boxplot1-exercise, fig.align='center', fig.height=4.5, fig.width=4.5, message=FALSE, exercise=TRUE, exercise.eval=TRUE, warning=FALSE}
## Load dataset
data(Loblolly)

## View the first few observations
head(Loblolly)
```

```{r boxplot1-exercise-solution, exercise.reveal_solution = TRUE}
## Boxplot in base R
boxplot(height ~ Seed, data = Loblolly)

## gglot2
ggplot(data = Loblolly, aes(x = Seed, y = height)) +
  geom_boxplot(fill = "grey") +
  theme_test()
```
<br>

2.3 Create a grouped boxplot of the pre-loaded dataset `insects` containing the lengths of an insect species across seven sites (labeled A to G) as a function of an invasive predator exclusion treatment (control = no predator exclusion, exclusion = predator exclusion). 'Grouped' means that multiple (nested) grouping variables are present in the data. The categories of the overarching grouping variable are indicated by the tick marks on the _x_-axis. The boxes of the nested grouping variable are shown side-by-side at each tick mark (i.e. for each level of the overarching grouping variable). In this example the site identity (`site`) represents the overarching grouping variable and the predator exclusion treatment (`treat`) indicates the nested grouping variable.

**Hint:** The `boxplot` function in base R accepts multiple grouping variables separated by `+` signs, _e.g._ `boxplot(y ~ groupvar1 + groupvar2, data = ...)`.

```{r boxplot2-exercise, exercise=TRUE, exercise.eval=TRUE, fig.align='center', fig.width=4.5, fig.height=4.5, warning=FALSE, message=FALSE}
## View the first few observations
head(insects)
```

```{r boxplot2-exercise-solution, exercise.reveal_solution = TRUE}
## Boxplot in base R
## Pre-define box positions
positions <- rep(1:7, each = 2) + rep(c(-0.15, 0.15), times = 7)

boxplot(length ~ treat + site, data = insects, col = c("white", "grey"), 
        width = rep(0.1, 14), at = positions, boxwex = 0.3) # NOTE: treat needs to come before site in the right-hand side of the equation to get the grouping correct. See help file for 'bxp' for an explanation of the graphical boxplot parameters that can be tweaked, e.g. add pars = list(whisklty = 1, staplewex = 0) to the boxplot code above.

## gglot2
ggplot(data = insects, aes(x = site, y = length, fill = treat)) +
  scale_fill_manual(values = c("white", "grey")) +
  geom_boxplot()
```
<br>

### Exercise 3 - Multi-panel figures

3.1 The built-in dataset `CO2` contains photosynthetic carbon dioxide uptake rates (µmol m^-2^ s^-1^) of a grass species in response to a chilling treatment (nonchilled vs. chilled conditions) and ambient CO~2~ concentration (mL L^-1^; see help file for `CO2`). Two provenances of this species were used in this trial, one from Quebec and another one from Mississippi (variable `Type`). Create a two-panel scatter plot figure showing each provenance in a separate panel and colour-code the treatment (nonchilled = white, chilled = black). Note that the control should always be displayed in the less conspicuous colour. Use the units given above in the axis titles.

```{r mult1-exercise, exercise=TRUE, exercise.eval=TRUE, fig.align='center', fig.width=6, fig.height=3, message=FALSE, warning=FALSE}
## Load data
data(CO2)

## View the first few observations
head(CO2)
```

```{r mult1-exercise-solution, exercise.reveal_solution = TRUE, fig.align='center', fig.width=6, fig.height=3.5}
## Traditional graphics
layout(rbind(c(1, 0, 2)), widths = c(1, lcm(0.3), 1))
# layout.show(n = 2) # check out the layout
op <- par(mar = rep(0, times = 4), oma = c(5, 4, 2, 1))

## Quebec provenance
plot(uptake ~ conc, data = CO2[CO2$Type == "Quebec", ], xlim = c(0, 1100), xaxs = "i", ylim = c(0, 50), yaxs = "i", ann = F, tcl = 0.35, mgp = c(3, 0.4, 0), las = 1, pch = 21, bg = c("white", "black")[CO2$Treatment], cex = 1.2, cex.axis = 1.1)

## Axis titles
mtext(text = expression(paste("CO"[2], " uptake (", mu, "mol m"^-2, " s"^-1, ")")), side = 2, line = 2, cex = 0.8)
mtext(text = expression(paste("Ambient CO"[2], " concentration (mL L"^-1, ")")), side = 1, line = 2.2, cex = 0.8, at = 1125)

## Provenance annotation
mtext(text = "Quebec", side = 3, line = 0.4, cex = 0.8)

## Legend
legend(x = 650, y = 15, legend = c("Nonchilled", "Chilled"), pch = 21, pt.bg = c("white", "black"), pt.cex = 1.2, bty = "n", cex = 1.1)

## Mississippi provenance
plot(uptake ~ conc, data = CO2[CO2$Type == "Mississippi", ], xlim = c(0, 1100), xaxs = "i", ylim = c(0, 50), yaxs = "i", ann = F, pch = 21, bg = c("white", "black")[CO2$Treatment], axes = F, cex = 1.2)
axis(1, tcl = 0.35, mgp = c(3, 0.4, 0), cex.axis = 1.1)
axis(2, labels = F, tcl = 0.35)
box()
## Provenance annotation
mtext(text = "Mississippi", side = 3, line = 0.4, cex = 0.8)
par(op)

## ggplot2
ggplot(data = CO2, aes(x = conc, y = uptake, fill = Treatment)) +
  geom_point(shape = 21) + # shape = 21 means filled circles
  labs(y = expression(paste("CO"[2], " uptake (", mu, "mol m"^-2, " s"^-1, ")")), x = expression(paste("Ambient CO"[2], " concentration (mL L"^-1, ")"))) +
  scale_fill_manual(labels = c("Nonchilled", "Chilled"), values = c("white", "black")) + # legend entries should be capitalised
  facet_wrap(facets = ~ Type) +
  theme_test()
```
<br>

### Exercise 4 - Figures with inset

We will use the built-in `faithful` dataset, which contains the waiting (variable `waiting`) between eruptions and the duration of the eruption events (variable `eruptions`) for the Old Faithful geyser in the Yellowstone National Park, USA. Plot the waiting time (minutes) as a function of the eruption duration (minutes) and add an histogram inset in the upper left-hand corner of the plot using both base R and _ggplot2_. There are plenty of _ggplot2_ extension packages for adding insets. We recommend the user-friendly _cowplot_ package but the choice is up to the reader. 

**Hint:** To enable a snug fit of the inset in base R, insert this code snippet `op <- par(mar = rep(0, times = 4), oma = c(5, 4, 2, 1))` before your plotting code to remove the figure margins and add an outer margin for the axis titles. At the end of your plotting code, insert `par(op)` to reset the default graphical parameter settings. 

```{r inset-exercise, exercise=TRUE, exercise.eval=TRUE, fig.align='center', fig.width=4.5, fig.height=4, message=FALSE, warning=FALSE}
## Load the data
data(faithful)

## View the first few observations
head(faithful)
```

```{r inset-exercise-solution, exercise.reveal_solution = TRUE, message=FALSE}
## Traditional graphics (base R)
## Create the layout
layout(rbind(c(1, 1, 1),
             c(1, 2, 1),
             c(1, 1, 1)), widths = c(0.15, 0.25, 0.6), heights = c(0.05, 0.25, 0.7)) # 1 is the main plot which comes first in the code below, 2 indicates the inset whose code comes second.

# layout.show(2) # check layout, needs to be outcommented before saving the figure

op <- par(mar = rep(0, times = 4), oma = c(5, 4, 2, 1)) # remove figure margins to create a snug fit of the inset and add outer margins to the overall figure to make room for axis titles
## Main plot
plot(waiting ~ eruptions, data = faithful, tcl = -0.3, mgp = c(2, 0.5, 0), las = 1, cex = 1.2)
mtext("Waiting time (min)", side = 2, line = 1.9, cex = 0.8)
mtext("Eruption time (min)", side = 1, line = 1.8, cex = 0.8)

## Inset
hist(faithful$waiting, main = NA, cex.axis = 0.8, tcl = -0.3, mgp = c(2, 0.5, 0), las = 1)
mtext("Frequency", side = 2, line = 1.6, cex = 0.6)
mtext("Waiting time (min)", side = 1, line = 1.4, cex = 0.6)
# box() # optional bounding box around the histogram
par(op) # reset default graphical parameters

## ggplot2
main <- ggplot(data = faithful, aes(x = eruptions, y = waiting)) +
  geom_point() +
  labs(y = "Waiting time (min)", x = "Eruption time (min)") +
  theme_test()
  
inset <- ggplot(data = faithful, aes(waiting)) +
  geom_histogram() +
  labs(y = "Frequency", x = "Waiting time (min)") +
  theme_test(base_size = 8)

## Combine the two plots
 ggdraw() +
  draw_plot(main) +
  draw_plot(inset, x = 0.15, y = 0.65, width = 0.3, height = 0.3)
```
